Phylogram for testing neutral evolution of a trait, chronogram to model selection.

### 2025-02-28 manuscript/main.tex
Update manuscript

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}
%TC:ignore
\maketitle

% Abstract (≤ 300 words)
\begin{abstract}
    The evolution of traits across different species is driven by a combination of selective and neutral processes, while their relative importance is subject to numerous debates.
    To disentangle them, different scenarios of evolution are modelled and compared.
    Typically, a neutrally evolving trait is modelled as changing randomly along the different branches of the species tree.
    This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time.
    Theoretically, changes for a neutral trait depend on the number of generations passed, which in turn correlates with time, hence the use of chronogram.
    But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases.
    Instead, for a phylogram the branch lengths represent sequence divergence (i.e. the number of substitutions), which absorb the effect of changing generation time and mutation rates.
    In this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram.
    However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions.
    Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift.
    Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

\keywords{Trait evolution \and Phylogenetics \and Phylogenetic Comparative Methods \and Chronogram \and Sequence divergence \and Neutral evolution}

@@

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
Across species, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed variation.
Such regimes of evolution are typically compared using phylogenetic comparative methods, where a continuous traits is modeled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
This modeling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.

% This comes from the fact that we use chronogram, and not phylogram.
% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}).
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
Instead, on a phylogram, the branch lengths can represent another quantity than time.
For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
Such a phylogram in unit of nucleotide divergence obtained from neutrally evolving genomic loci would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate the branch lengths of chronograms.

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_rational}
    \caption{
        Panel~A: Across many species, the evolution of continuous traits can be modeled as a stochastic process evolving along the branches of a phylogenetic tree.
        The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
        Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
        Instead, changes of a neutral trait should be better predicted a phylogram.
    }
    \label{fig:methods}
\end{figure*}
%TC:endignore

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes, by evaluating by the fit of a BM on both trees (Fig.~\ref{fig:methods}B).
In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
Altogether, by using genomic information we seek to provide additional approaches to model trait evolution.

\section*{Methods}
\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, multiple optima, moving optimum).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summry, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

@@

The phylogram is obtained from the same simulation setting on a set of independent 30.000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

\subsection*{Empirical dataset}
We analyzed a dataset of body and brain masses from mammals.
The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
We also extracted brain and body masses from the \textit{COMBINE} dataset~\citep{soria_combine_2021}.
The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
More specifically, the phylogram in unit of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood smoothing parameter of 1) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

@@

Finally, we implemented a model in \textit{RevBayes} containing a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length.
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favored over the chronogram.

@@

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process (see Methods, BM section).
To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) evolving under multiple optima, and 3) evolving under a moving optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
In such a case, the support is the same for both a phylogram and a chronogram (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree mimicking a mammalian range of changes.
In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).

%TC:ignore
\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_simulations}\\
    \vspace{10pt}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{D: simulations with constant generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{E: simulations with fluctuating generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
    \end{minipage}
    \caption{
        Panel~A: Wright-Fisher simulations with mutation, selection and drift.
        For a given individual, its genotypic value is encoded by a finite number of independent loci (meaning no linkage) contributing additively.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Panel~B: Drift is modeled by the resampling of parents.
        Selection is modeled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
        Panel~C: Example simulation of a trait evolving along a phylogenetic tree.
        Panels D \& E: Violinplot of posterior probability of the phylogram being favored over the chronogram for the different simulated regime of selection.
        Horizontal lines inside the violin show the result of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
        Results of simulations with constant generation time (Panel~D) and with fluctuating generation time, mutation rate and effective population size (Panel~E).
    }
    \label{fig:results-brownian}
\end{figure*}
%TC:endignore

>>>>

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}
%TC:ignore

\maketitle

% Abstract (≤ 300 words)
\begin{abstract}
The evolution of traits across different species is driven by a combination of selective and neutral processes, while their relative importance is subject to numerous debates. To disentangle them, different scenarios of evolution are modelled and compared. Typically, a neutrally evolving trait is modelled as changing randomly along the different branches of the species tree. This species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time. Indeed, the time that has passed correlates with the changes for a neutral trait, through the direct proxy of the number of generations that occurred. But, we argue that since species generation time can also vary along the phylogenetic tree, chronogram induces biases. Instead, for a phylogram, the branch lengths represent sequence divergence (i.e. the number of substitutions) which absorb the effect of changing generation time. Passed on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a Brownian process has more support on a phylogram than on a chronogram. However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions. Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift. Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

\keywords{Comparative studies \and Morphological evolution \and Genetic drift \and Macroevolution \and Chronogram \and Phylogram}

@@

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
By observing the changes in trait along a lineage, it is possible to speculate on the regime of selection acting on it.
But without selection, a trait might still changes due to random genetic drift.
As such distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
At a larger scale, from the observed variations of traits across species, such regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian Motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
Or in other words, for each  branch of the tree the value at descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing the trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, when trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, such alternative models of evolution raise issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a change in the fitness peak~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which is interpreted as a trait tracking a moving optimum~\citep{hansen_translating_1996, latrille_detecting_2024}.
One special case a continuously changing optimum is again the BM, where the optimum is changing along the phylogenetic tree as a random walk~\citep{hansen_comparative_2008, hansen_three_2024, holstad_evolvability_2024}.
This modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a fluctuating optimum.

% This comes from the fact that we use chronogram, and not phylogram.
% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
One blind spot of this modelling framework is the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depends directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the generation time is constant.
But, since species generation time might vary along the phylogenetic tree, modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.
Instead, on a phylogram, the branch lengths can represent another quantity than time, which can be used as the backbone to model trait evolution.
For example, it is known that using phylograms in unit of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylogram can also be obtained in unit of nucleotide divergence, which is equivalent to measuring the number of nucleotide substitutions occurring along a branch.
Such a phylogram in unit of nucleotide divergence, if obtained from neutrally evolving genomic loci, would absorb the effect of changing generation time and mutation rates, and thus would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Conversely, for a trait evolving under selection, using a chronogram would be more accurate since changes in the optimum fitness peak is extrinsic to the species and thus should be better correlated to time rather than the number of substitutions that occurred along a branch~\citep{hansen_comparative_2008}.
% Finally, from a statistical perspective, using nucleotide divergence would also remove the potential effect of model assumptions required to calibrate ancestral node ages (e.g. molecular clocks) to estimate the branch lengths of chronograms.

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_rational}
    \caption{
        Panel~A: Across many species, the evolution of continuous traits can be modelled as a stochastic process evolving along the branches of a phylogenetic tree.
        The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
        Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
        Instead, changes of a neutral trait should be better predicted a phylogram.
    }
    \label{fig:methods}
\end{figure*}
%TC:endignore

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if it is more accurate to use a phylogram (in unit of nucleotide divergence for neutral loci) instead of a chronogram as the underlying tree to model neutral trait changes (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
In other words, we seek to test if change of a neutral trait is better predicted a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum is better predicted by a chronogram.
Altogether, by using genomic information we seek to provide additional approaches to model trait evolution.

\section*{Methods}
\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape result in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch, if a switch occurs the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

@@

The phylogram is obtained from the same simulation setting on a set of independent 30.000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

\subsection*{Empirical dataset}
We analysed a dataset of body and brain masses from mammals.
The log-transformed values of body and brain masses were taken from \citet{tsuboi_breakdown_2018}.
We removed individuals not marked as adults and split the data into males and females due to sexual dimorphism in body and brain masses.
We also extracted brain and body masses from the \textit{COMBINE} dataset~\citep{soria_combine_2021}.
The mammalian genomic data are gathered from the Zoonomia project~\citep{genereux_comparative_2020}.
More specifically, the phylogram in unit of nucleotide divergence is estimated on a set of neutral markers in \citet{foley_genomic_2023}.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

@@

Finally, we implemented a model in \textit{RevBayes} containing a switch variable, denoted $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions.
Mathematically, $\PhyloSupport$ is a Bernoulli random variable with a prior probability of $0.5$.
If $\PhyloSupport = 0$, the branch lengths are those of the chronogram, and if $\PhyloSupport = 1$, the branch lengths are those of the phylogram.
The tree topology is fixed (same for both trees), and both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
As in the previous section, a Brownian process is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1000 gen., running of 10000 gen., 2 chains) is the probability the phylogram is favoured over the chronogram.

@@

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_simulations}\\
    \vspace{10pt}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{D: simulations with constant generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_cstGT_phylogram_support}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\small  \textbf{E: simulations with fluctuating generation time}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_phylogram_support}
    \end{minipage}
    \caption{
        Panel~A: Wright-Fisher simulations with mutation, selection and drift.
        For a given individual, its genotypic value is encoded by a finite number of independent loci (meaning no linkage) contributing additively.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Panel~B: Neutral evolution is modelled as a constant fitness regardless of phenotypic value.
        Selection is modelled by stabilizing selection around a optimum value, which is changing along the phylogenetic tree.
        Panel~C: Example simulation of a trait evolving along a phylogenetic tree.
        Panels D \& E: Violinplot of posterior probability of the phylogram being favoured over the chronogram ($\PhyloSupport$) across 100 replicate simulations for the different simulated regime of evolution.
        Horizontal lines inside violins are the estimated $\PhyloSupport$ of each replicate simulation.
        Regime of evolution: neutral regime (yellow), under a moving optimum (blue), and under multiple optima (red).
        Results of simulations with constant generation time (Panel~D) and with fluctuating generation time, mutation rate and effective population size (Panel~E).
    }
    \label{fig:results-brownian}
\end{figure*}
%TC:endignore

@@

\subsection*{Support for a phylogram over a chronogram}
% Simulations of a trait evolving neutrally versus under a moving optimum.
We developed a Bayesian model to estimate the support of phylogram over a chronogram for a trait evolving under a Brownian Motion (BM) process, denoted as $\PhyloSupport$, which range from 0 (full support for the chronogram) to 1.0  (full support for the phylogram), for more details see Methods (BM section).
To test this model, we simulated traits evolving on a phylogenetic trees under different regimes of selection: 1) neutrally evolving, 2) under a moving optimum and 3) under multiple optimum (see Methods, Fig.~\ref{fig:results-brownian}A-C).
As a control experiment we simulated with a constant generation time, mutation rate (per generation) and effective population size ($\Ne$).
In such a case, the support is the same for both a phylogram and a chronogram regardless of the regime of evolution (Fig.~\ref{fig:results-brownian}D), which is expected since the branch length are equivalent in both trees.
Next, we simulated changing generation time, mutation rate and $\Ne$ along the phylogenetic tree, mimicking a mammalian range of changes.
In this case, the BM was better fitted on a phylogram than on a chronogram for a neutral trait (Fig.~\ref{fig:results-brownian}E), with an average support for phylogram of $\MeanPhyloSupport = 0.94$ across $100$ replicate simulations, and $80 / 100$ replicates above the $0.95$ threshold.
Conversely, the fit of a BM on a chronogram was better than on a phylogram for a trait under selection, with an average support for phylogram of $\MeanPhyloSupport = 0.17$ for multiple optima ($69 / 100$ replicates below the $0.05$ threshold) and $\MeanPhyloSupport = 0.14$ for a moving optimum ($67 / 100$ below the $0.05$ threshold).
On the empirical mammalian dataset from \citet{tsuboi_breakdown_2018}, for body mass the support for the phylogram is $\PhyloSupport = 0.0$ when not splitting the males and females.
When splitting the dataset, the support for the phylogram is $\PhyloSupport_{\text{\Female}} = 0.85$ and $\PhyloSupport_{\text{\Male}} = 0.70$.
For the brain mass, the support for the phylogram is $\PhyloSupport = 0.0035$ on the mixed dataset, with $\PhyloSupport_{\text{\Female}} = 0.56$ and $\PhyloSupport_{\text{\Male}} = 0.51$.
On the \textit{COMBINE} dataset, which does not distinguish males and females, the support for the phylogram is $\PhyloSupport = 0.0$ for body mass and $\PhyloSupport = 0.0015$ for brain mass (Table~\ref{table:results-empirical}).
We also computed $\PhyloSupport$ using an approximate likelihood computation (REML), faster the larger trees, and found similar results (Table~\ref{table:results-empirical}).

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{A}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiBM}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{B}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiBM_NSwitch}
    \end{minipage}\\
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{C}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_OU}
    \end{minipage}
    \begin{minipage}{0.49\linewidth}
        \flushleft {\tiny  \textbf{D}}
        \includegraphics[width=\linewidth, page=1]{figures/simulation_varGT_MultiOU_NSwitch}
    \end{minipage}

    \caption{
        \textbf{Mis-classification of trait evolution on a phylogram.}
        Violinplot of posterior parameter estimates for different models of trait evolution and different simulated regime of selection.
        Horizontal lines inside the violin show the result of each replicate simulation.
        Simulations of 100 replicates per regime: trait evolving under a neutral regime (yellow) and under a moving optimum (blue).
        Wilcoxon rank-test are performed between the paired estimates on the phylogram and the chronogram.
        Panel~A \& B: Fit of a relaxed Brownian Motion (BM) with multiple rate parameters, using either a phylogram or a chronogram.
        Posterior estimate for the variance in rate parameters (Panel~A) and number of rate changes (Panel~B).
        Panel~C: Relative fit an Ornstein-Uhlenbeck (OU) process compared to BM, using either a phylogram or a chronogram.
        Posteriors estimates for the support of the OU process over the BM (reversible-jump).
        Panel~D: Fit of a relaxed OU with multiple optima, using either a phylogram or a chronogram.
        Posterior estimate for the number of optimum changes.
    }
    \label{fig:results-alternative}
\end{figure*}
%TC:endignore

\subsection*{Trait evolution on a phylogram and on a chronogram}
% Mis-classification of trait evolution on a phylogram.
Beside support for the underlying tree, ancestral state reconstruction can also be improved by the use of a chronogram or a phylogram.
For a neutral trait, ancestral trait reconstruction was also more accurate on a phylogram (Fig.~\ref{fig:results-distance}A, $r^2=0.95$) than on a chronogram (Fig.~\ref{fig:results-distance}B, $r^2=0.85$).
In contrast, for a trait evolving under a moving optimum, ancestral trait reconstruction was less accurate on a phylogram (Fig.~\ref{fig:results-distance}C, $r^2=0.79$) than on a chronogram (Fig.~\ref{fig:results-distance}D, $r^2=0.96$).
Additionally, instead of testing the fit of a BM with a single rate, we also fitted a multi-rate BM process (see Methods).
When fitting a multi-rate BM process, the estimated variance of rate parameters ($v$) was lower on a phylogram than on a chronogram for a trait evolving neutrally (Fig.~\ref{fig:results-alternative}A, yellow violins, Wilcoxon paired test with $p_{\text{value}}=8.8\times 10^{-14}$), and the number of rate shifts ($n$) was reflecting the prior on the phylogram while not on the chronogram, showing a bias (Fig.~\ref{fig:results-alternative}B, yellow violins).
Conversely, for a trait under a moving optimum, $v$ was higher on a phylogram than on a chronogram (Fig.~\ref{fig:results-alternative}A, blue violins, Wilcoxon paired test with  $p_{\text{value}}=8.9\times 10^{-15}$), and $n$ was reflecting the prior on the chronogram while not on the phylogram, showing a bias (Fig.~\ref{fig:results-alternative}B, blue violins).
Altogether, fitting a BM on a phylogram is more accurate for a trait evolving neutrally, but results in less accurate estimates for a trait under selection.
